Notes about Spin and TR Symmetry

### 2016-10-09 Notes about TR symmetry/Notes about TR symmetry.tex
Updating TR Symmetry.

% Following is for theorems etc environments
% http://tex.stackexchange.com/questions/45817/theorem-definition-lemma-problem-numbering && https://en.wikibooks.org/wiki/LaTeX/Theorems
\usepackage{amsthm}
\newtheorem{defi}{Definition}[section]
\newtheorem{thm}{Theorem}[section]
\newtheorem{lemma}{Lemma}[section]
\newtheorem{remark}{Remark}[section]
\newtheorem{prop}{Proposition}[section]
\newtheorem{coro}{Corollary}[section]
\theoremstyle{definition}
\newtheorem{ex}{Example}[section]

@@

    \subsection{From classical notion to property of TR symmetry}
    \label{sec:From-classical-notion-to-property-of-TR-symmetry}
    To get the time reversal operator, we can first look at the following
    equality (let $\Theta$ dentoes the time reversal operator):
    \begin{align}
        \left(1-\frac{iH}{\hbar}\delta t\right) \Theta \ket{\alpha}
        = \Theta\left(1-\frac{iH}{\hbar}(-\delta t)\right)\ket{\alpha}
        \label{eq:TRoper.fromClas2Quam.phyPic}
    \end{align}
    This follows from the following notion: a time reversed state is
    equivalent to the original state at an earlier time. Recall that
    $\left(1-\frac{iH}{\hbar}\delta t\right)$ produces an infinitesimal
    time translation.

    From equality \ref{eq:TRoper.fromClas2Quam.phyPic}, one finds:
    \begin{align}
        -iH\Theta = \Theta i H
    \end{align}

    One can show that \textbf{$\Theta$ must be antilinear}. Because if one
    assumes that $\Theta$ is linear, one will find that the energy
    spectrum for time reversal symmetric system will always has symmetric
    energy spectrum (i.e. with both and symmetric positive and negative
    energy), which is not true.

    So we have:
    \begin{itemize}
        \item $\Theta$ is antiunitary.
        \item $\Theta H=H\Theta$.
    \end{itemize}

    Recall that $\Theta= UK$, and $K$ is highly dependent on the basis
    chosen. So $U$ will depend on the basis too. To get $\Theta$ in a
    basis $\ket{e_i}$, obviously we only need to consider the result of
    $K\ket{e_i}$. This however, is hard to determine. However we can argue
    that in case when the basis has some classical connection, then by the
    Correspondance Principle we have, for example:
    \begin{align}
        \Theta \hat{p} \Theta^{-1}&= -\hat{p} \\
        \Theta \hat{x} \Theta^{-1}&= \hat{x} \\
        \Theta \mathbf{J} \Theta^{-1}&= -\mathbf{J}
    \end{align}
    with these one can easily find:
    \begin{align}
        \Theta \ket{p} &= \ket{-p} \\
        \Theta \ket{x} &= \ket{x} \\
        [x_i, p_j]\Theta \ket{} &= i\hbar \delta_{ij} \Theta\ket{}
    \end{align}
    where $\ket{}$ stands for any ket. Then clearly, $\Theta$ is just
    complex conjugation for wave function $\Psi(x)$. But $\Theta$ acts on
    $\Psi(p)$ will produces $\Psi^*(-p)$. The case for eigenstate of
    spin is a little complicated, which will be discussed in section
    \ref{sec:Time-reversal-operator-in-spin-j}.

    Following is a useful theorem, which is easy to proof (proof can be
    found on page 294 of \cite{sakurai}):
    \begin{thm}
        When Hamiltonian is invariant under time reversal and the energy
        eigenket is nondegenerate; then the corresponding energy
        eigenfunction is real (or, more generally, a real function times a
        phase factor independent of $x$)
    \end{thm}
    Therefore, a complex eigenfunction almost means a degenrate state, or
    the system is not invariant under time reversal.

    \subsection{Time reversal operator in spin \texorpdfstring{$j$}{}}
    \label{sec:Time-reversal-operator-in-spin-j}
    Using my guess \ref{eq:revSpin.getMatRot.Sn.Snket}:
    \begin{align*}
        \ket{\hat{\mathbf{n}};+}= e^{-iS_z \alpha/\hbar}
        e^{-iS_y \beta/\hbar} \ket{+}
    \end{align*}
    and by similar argument in that section \ref{sec:Eigenstate_for_Sn},
    we can argue that the state with opposite direction is
    \begin{align}
        \ket{\hat{\mathbf{n}};-}= e^{-iS_z \alpha/\hbar}
        e^{-iS_y (\beta+\pi)/\hbar} \ket{+}
    \end{align}
    Again, I am using notation $\ket{-}$ to denotes $\ket{j,-j}$.

    Since time reversal reverses the angular momentum, I expects:
    \begin{align}
        \Theta \ket{\hat{\mathbf{n}};+} = \eta \ket{\hat{\mathbf{n}};-}
    \end{align}
    where $\eta$ is just some unnecessay factor.

    With above relationship, one can easily see that:
    \begin{align}
        \Theta = \eta e^{-i\pi J_y/\hbar}K
    \end{align}
    The proof is trivial for $j=\frac{1}{2}$, but I cannot find a proof
    for other values of $j$.

    It is also interesting to note that in general:
    \begin{align}
        \Theta^2 &= -1 \text{ if $j$ is a half-integer} \\
        \Theta^2 &= 1 \text{ if $j$ is an integer}
    \end{align}
    However, this can only be proved when $j=\frac{1}{2}$, in which case
    the proof is also straightforward if we use the formula mentioned in
    section \ref{sec:About-exp(-i*pi-S_y/hbar)}.

    However, if we imagine a spin $j$ system as composed all just of
    electrons of spin $\frac{1}{2}$, then we have a strong argument that
    the above formula is true.
    \subsection{Kramer's theorem and the lifting of degenracy}

\section{Anchor}
\label{sec:Anchor}

\begin{thebibliography}{1}
    \bibitem{sakurai} Modern Quantum Mechanics. J.J. Sakurai.
    \bibitem{bernevig} Introduction to Topological Insulators. Bernevig.
    \bibitem{math.SE-adjoint-of-antilinear}\href{http://physics.stackexchange.com/questions/45227/existence-of-adjoint-of-an-antilinear-operator-time-reversal}{Existence of adjoint of an antilinear operator, time reversal}
\end{thebibliography}
\printnomenclature
\section{License}
The entire content of this work (including the source code
for TeX files and the generated PDF documents) by 
Hongxiang Chen (nicknamed we.taper, or just Taper) is
licensed under a 
\href{http://creativecommons.org/licenses/by-nc-sa/4.0/}{Creative 
Commons Attribution-NonCommercial-ShareAlike 4.0 International 
License}. Permissions beyond the scope of this 
license may be available at \url{mailto:we.taper[at]gmail[dot]com}.
\end{document}

>>>>

% Following is for theorems etc environments
% http://tex.stackexchange.com/questions/45817/theorem-definition-lemma-problem-numbering && https://en.wikibooks.org/wiki/LaTeX/Theorems
\usepackage{amsthm}
\newtheorem{defi}{Definition}[section]
\newtheorem{thm}{Theorem}[section]
\newtheorem{lemma}{Lemma}[section]
\newtheorem{remark}{Remark}[section]
\newtheorem{prop}{Proposition}[section]
\newtheorem{coro}{Corollary}[section]
\newtheorem{fact}{Fact}[section]
\theoremstyle{definition}
\newtheorem{ex}{Example}[section]
\newtheorem{argument}{Argument}[section]

@@

    \subsection{From classical notion to properties of TR symmetry}
    \label{sec:From-classical-notion-to-property-of-TR-symmetry}
    To get the time reversal operator, we can first look at the following
    equality (let $T$ dentoes the time reversal operator):
    \begin{align}
        \left(1-\frac{iH}{\hbar}\delta t\right) T \ket{\alpha}
        = T\left(1-\frac{iH}{\hbar}(-\delta t)\right)\ket{\alpha}
        \label{eq:TRoper.fromClas2Quam.phyPic}
    \end{align}
    This follows from the following notion: a time reversed state is
    equivalent to the original state at an earlier time. Recall that
    $\left(1-\frac{iH}{\hbar}\delta t\right)$ produces an infinitesimal
    time translation.

    From equality \ref{eq:TRoper.fromClas2Quam.phyPic}, one finds:
    \begin{align}
        -iHT = T i H
    \end{align}

    One can show that \textbf{$T$ must be antilinear}. Because if one
    assumes that $T$ is linear, one will find that the energy
    spectrum for time reversal symmetric system will always has symmetric
    energy spectrum (i.e. with both and symmetric positive and negative
    energy), which is not true.

    So we have:
    \begin{itemize}
        \item $T$ is antiunitary.
        \item $T H=HT$ for a time reversal symmetric state.
    \end{itemize}

    Recall that $T= UK$, and $K$ is highly dependent on the basis
    chosen. So $U$ will depend on the basis too. To get $T$ in a
    basis $\ket{e_i}$, obviously we only need to consider the result of
    $K\ket{e_i}$. This however, is hard to determine. However we can argue
    that in case when the basis has some classical connection, then by the
    Correspondance Principle we have, for example:
    \begin{align}
        T \hat{p} T^{-1}&= -\hat{p} \\
        T \hat{x} T^{-1}&= \hat{x} \\
        T \mathbf{J} T^{-1}&= -\mathbf{J}
    \end{align}
    with these one can easily find:
    \begin{align}
        T \ket{p} &= \ket{-p} \\
        T \ket{x} &= \ket{x} \\
        [x_i, p_j]T \ket{} &= i\hbar \delta_{ij} T\ket{}
    \end{align}
    where $\ket{}$ stands for any ket. Then clearly, $T$ is just
    complex conjugation for wave function $\Psi(x)$. But $T$ acts on
    $\Psi(p)$ will produces $\Psi^*(-p)$. The case for eigenstate of
    spin is a little complicated, which will be discussed in section
    \ref{sec:Time-reversal-operator-in-spin-j}.

    \subsection{Properties of TR operator}
    \label{sec:Properties-of-TR-operator}

    Following is a useful theorem, which is easy to proof (proof can be
    found on page 294 of \cite{sakurai}):
    \begin{thm}
        \label{thm:TRoper.fromCM2QM.TR_nondegen_real}
        When Hamiltonian is invariant under time reversal and the energy
        eigenket is nondegenerate; then the corresponding energy
        eigenfunction is real (or, more generally, a real function times a
        phase factor independent of $x$)
    \end{thm}
    Therefore, a complex eigenfunction almost means a degenrate state, or
    the system is not invariant under time reversal.

    The next theorem is about $T^2$. Classically, since there is only
    two directions in time, we expect $T^2=1$. But quantum states are
    rays, so there is the freedom of phase. Then $T^2 = \phi$ for
    some $|\phi|^2=1$. But better, we have:

    \begin{lemma}
        $T^2 = \pm \mathbbm{1}$
    \end{lemma}
    \begin{proof}
        Proof can be found on page 34 of \cite{bernevig}. But the proof is
        so clever that I want to include it here. First observe that 
        $U U^\dagger = \mathbbm{1}$, implies that $U^* U^T = \mathbbm{1}$
        by transposing both sides. Next one use the expression $T=UK$ to
        expand $T^2$, only to find that $U=\phi U^T$, this means that
        $\phi^2=1$. Hence $\phi=\pm 1$.
    \end{proof}

    We then have the useful fact:
    \begin{fact}
        For any kets $\ket{\alpha},\ket{\beta}$, we have
        \begin{align}
            \pm \braket{\alpha|T\beta} = 
                \braket{\beta|T\alpha}
        \end{align}
        The sign is the same as in $T^2=\pm 1$. Specifically, one has
        \begin{align}
            \braket{\alpha |T \alpha} = 0
        \end{align}
        when $T^2= -\mathbbm{1}$.
    \end{fact}
    \begin{proof}
        This is simple by using 
        $\braket{T\alpha|T\beta} = \braket{\beta|\alpha}$, 
        and replacing $\ket{\alpha}$ with $\ket{T \alpha}$.
    \end{proof}

    \subsection{Time reversal operator in system with spin 
    \texorpdfstring{$j$}{}}
    \label{sec:Time-reversal-operator-in-spin-j}
    Using my guess \ref{eq:revSpin.getMatRot.Sn.Snket}:
    \begin{align*}
        \ket{\hat{\mathbf{n}};+}= e^{-iS_z \alpha/\hbar}
        e^{-iS_y \beta/\hbar} \ket{+}
    \end{align*}
    and by similar argument in that section \ref{sec:Eigenstate_for_Sn},
    we can argue that the state with opposite direction is
    \begin{align}
        \ket{\hat{\mathbf{n}};-}= e^{-iS_z \alpha/\hbar}
        e^{-iS_y (\beta+\pi)/\hbar} \ket{+}
    \end{align}
    Again, I am using notation $\ket{-}$ to denotes $\ket{j,-j}$.

    Since time reversal reverses the angular momentum, I expects:
    \begin{align}
        T \ket{\hat{\mathbf{n}};+} = \eta \ket{\hat{\mathbf{n}};-}
    \end{align}
    where $\eta$ is just some unnecessay factor.

    With above relationship, one can easily see that:
    \begin{align}
        T = \eta e^{-i\pi J_y/\hbar}K
    \end{align}
    The proof is trivial for $j=\frac{1}{2}$, but I cannot find a proof
    for other values of $j$.

    It is also interesting to note that in general:
    \begin{thm} \begin{align}
        \label{eq:TRoper.TRoper.square_minus}
        T^2 &= -1 \text{ if $j$ is a half-integer} \\
        T^2 &= 1 \text{ if $j$ is an integer}
        \label{eq:TRoper.TRoper.square_plus}
    \end{align} \end{thm}
    However, this can only be proved when $j=\frac{1}{2}$, in which case
    the proof is also straightforward if we use the formula mentioned in
    section \ref{sec:About-exp(-i*pi-S_y/hbar)}.

    However, if we imagine a spin $j$ system as composed all just of
    electrons of spin $\frac{1}{2}$, then we have a strong argument that
    the above formula is true.
    \subsection{Kramer's theorem and the lifting of degenracy}
    \label{sec:Kramers-theorem-and-the-lifting-of-degenracy}
    When one is concerned with the energy spectrum, one encounters the
    Kramer's theorem, which says:
    \begin{thm}[Kramer's theorem]
        For a system with halt-integer $j$ spin, the degree of degeneracy
        is at least two.
    \end{thm}
    \begin{proof}
        Let $\ket{n}$ be an energy eigenstate. If the system is time
        reversal symmetric, i.e. $[H,T]=0$, then $T \ket{n}$ is
        another energy eigenstate with the same energy. Supposing there is
        no degeneracy, then we have
        $$ \ket{n} = e^{i\eta}T\ket{n}$$
        for some $\eta\in \mathbb{R}$. Apply $T$ on both sides, then
        we have
        \begin{align*}
            e^{-i\eta} T^2 \ket{n} &= \pm e^{-i\eta}\ket{n} \\
            &= T \ket{n} = e^{i\eta} \ket{n}
        \end{align*}
        Clearly this is a contracdiction unless $T^2=1$, which is
        true only for interger spin system by equation
        \ref{eq:TRoper.TRoper.square_plus} and
        \ref{eq:TRoper.TRoper.square_minus}.
    \end{proof}
    Another way to show this is presented in page 37 of
    \cite{bernevig}. That method is too troublesome and basis dependent 
    \footnote{Borrowing the jargon of mathematicians, not canonical,
        geometric, or intrinsic.}
    to be shown here.
    But it shows an interesting fact that $T^2=-1$, actually
    implies that $U$, its unitary component mentioned in section
    \ref{sec:Property-of-symmetry-operator}, is antisymmetric. This is
    obvious when one juxtapose the following two facts:
    $$ U (U^*)^T = \mathbbm{1}$$
    $$ T^2 = UKUK = U U^* = -\mathbbm{1} $$

    By Kramer's theorem, one also says that for system with half-integer
    spin, one \textit{lift the degenracy by breaking the time reversal
    symmetry of the system.}

        \subsubsection{Argument against Bernevig's result}
        \label{sec:Argument again Bernevig's result}

        An additional point by Bernevig is about the transition probability
        $\bra{T \psi}H\ket{\psi}$ from a state to its time reversed
        state.

        Here I first mention two points:
        \begin{enumerate}
            \item When $T H=HT$, and $T=UK$, one can deduce
                $UH^* = HU$, or $U^*H=H^*U^*$, i.e.
                $(U^*)_{mp}H_{pn}=(H^*)_{mp}(U^*)_{pn}$.
            \item As mentioned before in section
                \ref{sec:Kramers-theorem-and-the-lifting-of-degenracy},
                $U$ is antisymmetric, so $U^*$ is antisymmetric too.
        \end{enumerate}

        Then begin Bernevig's wrong calculation (Summation implied when
        index is repeated twice):
        \begin{align}
            & \braket{T\psi|H|\psi} = (U_{mp}K\psi_p)^*H_{mn}\psi_n 
            \nonumber\\
            =& U_{mp}^* \psi_p H_{mn} \psi_n 
            \nonumber\\
            =& (U^\dagger)_{pm}\psi_p H_{mn}\psi_n 
            \label{eq:TRoper.Kramer.bernevig1}
            \\
            =& (U^\dagger)_{pm}\psi_p (THT^{-1})_{mn} \psi_n 
            \nonumber\\       
            =& (U^\dagger)_{pm}\psi_p \left(U_{mr}KH_{rq}(-U_{qn}K)\right)
                \psi_n
            \nonumber\\       
            =& - (U^\dagger)_{pm}\psi_p \left(U_{mr}H^*_{rq}U^*_{qn}\right)
                \psi_n
            \nonumber\\       
            =& - \psi_p \left((U^\dagger)_{pm}
                U_{mr}\right)H^*_{rq}U^*_{qn}\psi_n
            \nonumber\\       
            =& - \psi_p H^*_{pq}U^*_{qn}\psi_n
            \nonumber\\       
            \text{(by point 1)}=& - \psi_p U^*_{pq}H_{qn}\psi_n 
            \nonumber\\       
            \text{(by point 2)}=& U^*_{qp}\psi_p H_{qn} \psi_n 
            \nonumber\\       
            \text{(by the first line)}=& \braket{T\psi|H|\psi}
        \end{align}
        So I see that he just circles back to the origin...

        No I give an argument that statement in page 37 of
        \cite{bernevig} cannot be true.
        \begin{argument}
            Suppose we can prove that $\braket{T \psi|H|\psi}=0$.
            Then $\braket{\psi|HT |\psi}=0$, notice that $H=T
            HT^{-1}$, and $T^{2}=-1$, one can see that
            $\braket{\psi|H|\psi}=0$. Such is too strong an result. For
            example, if $\ket{\psi}$ is any eigenket of $H$, then our
            argument shows that its eigenvalue is zero. Does this means
            that $H$ is just the trivial operator $0$? Of course not!
            Therefore the Bernevig's result is wrong.
        \end{argument}

    \subsection{Time reversal symmetry in crystal}
    \label{sec:Time-reversal-symmetry-in-crystal}

        \subsubsection{Time reversal symmetry in spinless crystal}
        \label{sec:Time-reversal-symmetry-in-spinless-}
        Since there is not spin, time reversal symmetry does not alter a
        creation/annihilation of a particle. Then:
        \begin{align}
            Tc_j T^{-1} = c_j
        \end{align}
        Then, in the fourier expansion of $c_j$:
        \begin{align}
            c_j = \sum_k e^{ikR_j} c_k
        \end{align}
        Since $T$ acting on the left, will change $e^{ikR_j}$ into
        $e^{-ikR_j}$, then naturally:
        \begin{align}
            Tc_k T^{-1} = c_{-k}
        \end{align}
        Then, for a time reversal symmetric system, let 
        $H = \sum_k h(k)c_k^\dagger c_k$. Since $T H T^{-1}=H$,
        $T$ has to flip the sign of $k$ again, so:
        \begin{align}
            T h(k) T^{-1} = h(-k) \,\text{(For TR symmetric
            state)}
            \label{eq:TRoper.trInCry.trInSpinless.hk}
        \end{align}

        \paragraph{Implications for crystal systems} The above result has
        profound implications for crystal with zero spin.
        \begin{thm}
            For crystal with zero spin, one has in crystal momentum space,
            always at least a double degeneracy, except for certain point
            symmetric to $k$ and $-k$.
        \end{thm}
        \begin{proof}
            By equation \ref{eq:TRoper.trInCry.trInSpinless.hk}, one has
            $\phi(k)$ and $T\phi(k)$ corresponding to eigenstate in
            $k$ and $-k$. When $T$ acts on a wavefunction, its
            creation operator is affected and its ... (paused here).
        \end{proof}

\section{Anchor}
\label{sec:Anchor}
\begin{thebibliography}{1}
    \bibitem{sakurai} Modern Quantum Mechanics. J.J. Sakurai.
    \bibitem{bernevig} Introduction to Topological Insulators. Bernevig.
    \bibitem{math.SE-adjoint-of-antilinear}\href{http://physics.stackexchange.com/questions/45227/existence-of-adjoint-of-an-antilinear-operator-time-reversal}{Existence of adjoint of an antilinear operator, time reversal}
\end{thebibliography}
\printnomenclature
\section{License}
The entire content of this work (including the source code
for TeX files and the generated PDF documents) by 
Hongxiang Chen (nicknamed we.taper, or just Taper) is
licensed under a 
\href{http://creativecommons.org/licenses/by-nc-sa/4.0/}{Creative 
Commons Attribution-NonCommercial-ShareAlike 4.0 International 
License}. Permissions beyond the scope of this 
license may be available at \url{mailto:we.taper[at]gmail[dot]com}.
\end{document}
